---
title: "Imputation Pipeline: QUALITY CONTROL"
output: html_document
author: "ADV - ODAP - ICO"
date: "`r Sys.Date()`"
params:
  dataDir:
    label: "dataDir"
    value: ""
  plinkFile:
    label: "plinkFile:"
    value: ""
---

```{r setup, include=FALSE}

library(knitr)

dir.create("./QC/",showWarnings=FALSE)
outputDir<-"./QC/"

# read samples to mark them
fam <- read.table(paste0(dataDir, plinkFile, ".fam") ,stringsAsFactors = FALSE, header = FALSE, sep = "")
rownames(fam) <- fam[,2]

```

<br>

Filter out SNPs that don't pass quality control.

Mark samples that don't pass quality control.

<br>

#### Requisites:

* R 4.0.3

* plink 1.9

* AIMS_SNPs.txt file must be in the data folder

* Data must be in plink format (bed + bim + fam)

<br>  



### a) Missing SNPs:

Filter SNPs with 5% of missings

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# missing proportion
system(paste0("plink --bfile ", dataDir, plinkFile, " --missing --out ", outputDir, "missings"))
missings <- read.table(paste0(outputDir, "missings.lmiss"), header=TRUE)

## SNPs to exclude 
exclude_miss <- subset(missings, F_MISS > 0.05)
write.table(exclude_miss[,2], file = paste0(outputDir, "snp_exclude_miss.txt"), col.names = FALSE, row.names = FALSE, quote = FALSE, sep = "\t")

cat("\n")
cat(paste0("There are ",nrow(exclude_miss)," SNPs with more than 5% of missings. The complete list of SNPs to remove will be saved in SNPsmissings.txt"))
cat("\n")


## print plot
histplot <- hist(missings$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num SNPs", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.15, max(histplot$counts)/2, "removed SNPs", adj = 0, col = "darkred")
box(lwd = 2)

## save plot
png(paste0(outputDir, "SNPmissings.png"), res = 200, 1500, 1500)
histplot <- hist(missings$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num SNPs", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.15, max(histplot$counts)/2, "removed SNPs", adj = 0, col = "darkred")
box(lwd = 2)
dev.off()

# save excluded SNPs
write.table(exclude_miss, file = paste0(outputDir, "SNPsmissings.txt"), row.names = FALSE, quote = FALSE, sep = "\t")

```

<br>

### b) Missing samples: 

Mark samples with 10% of missings

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# missing proportion in samples
missInd <- read.table(paste0(outputDir, "missings.imiss"), header=TRUE)

## Samples to mark
exclude_miss <- subset(missInd, F_MISS > 0.1)

# filter SNPs with missings
system(paste0("plink --bfile ", dataDir, plinkFile, " --exclude ", outputDir, "snp_exclude_miss.txt --make-bed --out ", outputDir, plinkFile, "_filt"))

######################################################
# Mark which samples to exclude

fam$ExcludeMissings <- "No"
fam[as.character(exclude_miss[,2]),"ExcludeMissings"] <- "Yes"

cat("\n")
cat(paste0("There are ",sum(fam$ExcludeMissings=="Yes")," samples with more than 10% of missings."))
cat("\n")

if(sum(fam$ExcludeMissings=="Yes")>0) print(kable(fam[fam$ExcludeMissings=="Yes",c(1,2,7)],row.names=FALSE))

## print plot
histplot <- hist(missInd$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num samples", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.11, max(histplot$counts)/2, "marked samples", adj = 0, col = "darkred")
box(lwd = 2)

## save plot
png(paste0(outputDir, "SAMPLEmissings.png"), res = 200, 1500, 1500)
histplot <- hist(missInd$F_MISS, breaks = 20, xlab = "proportion of missings", ylab = "num samples", main = plinkFile)
abline(v = 0.05, col = "darkred", lwd = 2)
axis(1, 0.05, 0.05, col.axis = "darkred", cex.axis = 0.5)
arrows(0.05, max(histplot$counts)/2, 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(0.15, max(histplot$counts)/2, "marked samples", adj = 0, col = "darkred")
box(lwd = 2)
dev.off()

# remove some files
unlink(paste0(outputDir,"missings.imiss"))
unlink(paste0(outputDir,"missings.lmiss"))
unlink(paste0(outputDir,"missings.hh"))
unlink(paste0(outputDir,"missings.log"))
unlink(paste0(outputDir,"snp_exclude_miss.txt"))
unlink(paste0(outputDir,plinkFile,"_filt.log"))
unlink(paste0(outputDir,plinkFile,"_filt.hh"))

```

<br>

### c) Sex concordance:

Mark samples with no sex concordance

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# check sex concordance
system(paste0("plink --bfile ", outputDir, plinkFile, "_filt --check-sex --out ", outputDir, "sexConc"))
sex <- read.table(paste0(outputDir, "sexConc.sexcheck"), header = TRUE)

# filter in samples with no concordance
sex <- sex[sex$STATUS!="OK" & sex$SNPSEX!=0 & sex$PEDSEX!=0,]

# mark samples
fam$ExcludeSex <- "No"
fam[as.character(sex[,2]),"ExcludeSex"] <- "Yes"

cat("\n")
cat(paste0("There are ",sum(fam$ExcludeSex=="Yes")," samples with no sex concordance."))
cat("\n")

if(sum(fam$ExcludeSex=="Yes")>0) print(kable(fam[fam$ExcludeSex=="Yes",c(1,2,8)],row.names=FALSE))

# remove some files
unlink(paste0(outputDir,"sexConc.hh"))
unlink(paste0(outputDir,"sexConc.log"))

```

<br>

### d) Heterozigosity: 

Mark samples with high or low heterozigosity (mean+-4sd)

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# check heterozigosity
system(paste0("plink --bfile ", outputDir, plinkFile, "_filt --het --out ", outputDir, "heterozigotsity"))
hetero <- read.table(paste0(outputDir, "heterozigotsity.het"), header = TRUE)

# Heterozygosity rate
hetero$het <- (hetero[,5] - hetero[,3]) / hetero[,5]

# Cheking mean+-4sd
meanup <- mean(hetero$het) + 4 * sd(hetero$het)
meandown <- mean(hetero$het) - 4 * sd(hetero$het)
pos <- meandown < hetero$het & hetero$het < meanup

# samples to mark
exclude_het <- hetero[pos==FALSE,]

# mark samples
fam$ExcludeHeteroz <- "No"
fam[as.character(exclude_het[,2]),"ExcludeHeteroz"] <- "Yes"

cat("\n")
cat(paste0("There are ",sum(fam$ExcludeHeteroz=="Yes")," samples with high or low heterozigosity."))
cat("\n")

if(sum(fam$ExcludeHeteroz=="Yes")>0) print(kable(fam[fam$ExcludeHeteroz=="Yes",c(1,2,9)],row.names=FALSE))


## print plot
histplot <- hist(hetero$het, breaks = 20, xlab = "Heterozygosity rate", ylab = "num samples", main = plinkFile)
abline(v = meanup, col = "darkred", lwd = 2)
axis(1, meanup, round(meanup, 2), col.axis = "darkred", cex.axis = 0.5)
arrows(meanup, max(histplot$counts)/2, meanup + 0.01, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(meanup + 0.03, max(histplot$counts)/1.9, "marked samples", adj = 1, col = "darkred")

abline(v = meandown, col = "darkred", lwd = 2)
axis(1, meandown, round(meandown, 2), col.axis = "darkred", cex.axis = 0.5)
arrows(meandown, max(histplot$counts)/2, meandown - 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(meandown - 0.1, max(histplot$counts)/1.9, "marked samples", adj = 0, col = "darkred")
box(lwd = 2)


## save plot
png(paste0(outputDir, "Heterozigosity.png"), res = 200, 1500, 1500)
histplot <- hist(hetero$het, breaks = 20, xlab = "Heterozygosity rate", ylab = "num samples", main = plinkFile)

abline(v = meanup, col = "darkred", lwd = 2)
axis(1, meanup, round(meanup, 2), col.axis = "darkred", cex.axis = 0.5)
arrows(meanup, max(histplot$counts)/2, meanup + 0.01, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(meanup + 0.03, max(histplot$counts)/1.9, "marked samples", adj = 1, col = "darkred")

abline(v = meandown, col = "darkred", lwd = 2)
axis(1, meandown, round(meandown, 2), col.axis = "darkred", cex.axis = 0.5)
arrows(meandown, max(histplot$counts)/2, meandown - 0.1, max(histplot$counts)/2, length = 0.05, col = "darkred", code = 2, lwd = 2)
text(meandown - 0.1, max(histplot$counts)/1.9, "marked samples", adj = 0, col = "darkred")

box(lwd = 2)
dev.off()

# remove some files
unlink(paste0(outputDir,"heterozigotsity.hh"))
unlink(paste0(outputDir,"heterozigotsity.het"))
unlink(paste0(outputDir,"heterozigotsity.log"))

```

<br>

### e) Duplicates and relatedness:

Mark duplicated and related samples (pi_hat > 0.8)

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# check relatedness
system(paste0("plink --bfile ", outputDir, plinkFile, "_filt --genome --out  ", outputDir, "duplicates"))
dups <- read.table(paste0(outputDir, "duplicates.genome"), header = TRUE)

# filter in related or duplicated individuals
dups <- dups[dups$PI_HAT > 0.8,]

# mark samples
fam$ExcludeDups <- "No"

for(dp in 1:nrow(dups)){
  fam[as.character(dups[dp,2]),"ExcludeDups"] <- as.character(dups[dp,4])
  fam[as.character(dups[dp,4]),"ExcludeDups"] <- as.character(dups[dp,2])
}

cat("\n")
cat(paste0("There are ",sum(fam$ExcludeDups!="No")," duplicated and related samples."))
cat("\n")

if(sum(fam$ExcludeDups!="No")>0){
  print(kable(dups[,c(2,4,10)],row.names=FALSE))
  print(kable(fam[fam$ExcludeDups!="No",c(1,2,10)],row.names=FALSE))
} 


# remove some files
unlink(paste0(outputDir,"duplicates.log"))
unlink(paste0(outputDir,"duplicates.hh"))
unlink(paste0(outputDir,"duplicates.genome"))

```

<br>

### f) Hardy-Weinberg equilibrium:

Filter SNPs that have a HWE p-value < 1e-04

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# check HWE
system(paste0("plink --bfile ", outputDir, plinkFile, "_filt --hardy --out ", outputDir, "hwe"))
hwe <- read.table(paste0(outputDir, "hwe.hwe"), header = TRUE)
# hwe <- hwe[hwe$TEST=="UNAFF",] # uncomment if you have cases & controls 

## SNPs to exclude
exclude_hwe <- subset(hwe, P<1e-04)

cat("\n")
cat(paste0("There are ",nrow(exclude_hwe)," snps that don't pass the HWE.  The complete list of SNPs to remove will be saved in SNPsHWE.txt"))
cat("\n")

# Remove SNPs
write.table(exclude_hwe[,2],file = paste0(outputDir, "snp_exclude_hwe.txt"), quote = FALSE, col.names = FALSE, row.names = FALSE)
system(paste0("plink --bfile ", outputDir, plinkFile, "_filt --exclude ", outputDir, "snp_exclude_hwe.txt --make-bed --out ", outputDir, plinkFile, "_QC"))

# remove some files
unlink(paste0(outputDir,"snp_exclude_hwe.txt"))
unlink(paste0(outputDir,"hwe.hh"))
unlink(paste0(outputDir,"hwe.log"))
unlink(paste0(outputDir,"hwe.hwe"))
unlink(paste0(outputDir,plinkFile, "_filt.*"))
unlink(paste0(outputDir,plinkFile, "_QC.hh"))
unlink(paste0(outputDir,plinkFile, "_QC.log"))

# save removed SNPs 
write.table(exclude_hwe, file = paste0(outputDir, "SNPsHWE.txt"), row.names = FALSE, quote = FALSE, sep = "\t")


```

<br>

### g) Genetic population statification:

Mark samples outside the ancestry pupulation cluster of data using AIMS (Ancestry-informative marker SNPs) (abs(PC1) > 0.1 & abs(PC2) > 0.1). 

```{r, eval=TRUE, echo=FALSE, warning=FALSE, results='asis', error=FALSE}

# extract AIM SNPs
system(paste0("plink --bfile ", outputDir, plinkFile, "_QC --extract ", dataDir, "AIMS_SNPs.txt --make-bed --out  ", outputDir, plinkFile, "_AIMS"))

# PCA
system(paste0("plink --bfile ", outputDir, plinkFile, "_AIMS --pca --out  ", outputDir, "pca"))
pca <- read.table(paste0(outputDir, "pca.eigenvec"), header = FALSE)

## print plot
plot(pca[,3], pca[,4], pch = 20, xlab = "PCA 1", ylab = "PCA 2", main = plinkFile)

# save plot
png(paste0(outputDir, "Ancestry.png"), res = 200, 1500, 1500)
plot(pca[,3], pca[,4], pch = 20, xlab = "PCA 1", ylab = "PCA 2", main = plinkFile)
dev.off()

# remove some files
unlink(paste0(outputDir, plinkFile, "_AIMS.*"))
unlink(paste0(outputDir, "pca.log"))
unlink(paste0(outputDir, "pca.eigenval"))

# save marked samples
write.table(fam, file = paste0(outputDir, "marked_samples.txt"), sep = "\t", quote = FALSE, row.names = FALSE)

```


#### Output files:

The output files of the QC anlysis are:

* Three 'good quality' SNPs filtered in plink files (plinkFile_QC.bed; plinkFile_QC.bim and plinkFile_QC.fam)

* Two plots of missigness (SNPmissings.png and SAMPLEmissings.png)

* The list of SNPs with more than 5% of missings (SNPmissings.txt)

* Sex concordance file (sexConc.sexcheck)

* Heterozigosity plot (Heterozigosity.png)

* The list of SNPs not passing HWE (SNPsHWE.txt)

* The PCA components of AIM SNPs and the two first components plot (pca.eigenvec and Ancestry.png)

* A txt with the samples marked (marked_samples.txt). It is like a fam format file with some extra columns:

    + V1: Family ID ('FID')

    + V2: Within-family ID ('IID'; cannot be '0')

    + V3: Within-family ID of father ('0' if father isn't in dataset)

    + V4: Within-family ID of mother ('0' if mother isn't in dataset)

    + V5: Sex code ('1' = male, '2' = female, '0' = unknown)

    + V6: Phenotype value ('1' = control, '2' = case, '-9'/'0'/non-numeric = missing data if case/control)

    + ExcludeMissings: if 'Yes', sample has more than 10% of missings.

    + ExcludeSex: if 'Yes', sample hasn't sex concordance.

    + ExcludeHeteroz: if 'Yes', sample has a high heterozigosity.

    + ExcludeDups: if 'Yes', sample is duplicated or related to another sample.

<br>  
<br>  


